Six steps to a better relationship with your future self.

### 2011-03-25 workflow.Rnw
Almost done

This idea inspires a seminar paper in the spring. A conference paper
arises from the seminar paper in collaboration with another student in
2012. A dissertation chapter descends from the conference paper in
2013. Other dissertation chapters take up 2014. A submission to a
journal with the original co-author and a new collaborator happens in
2015. Revision and resubmission wait until 2017 while harried editors,
reviewers and authors strive to balance research, teaching, service,
and life. By now, the three lucky collaborators work as professors in
three different universities. In 2018 a child is born and a paper is
published. The United Nations takes an interest in the paper in 2019
and hosts a conference to discuss implications of the research. In
2020 a first year graduate student in a coffee shop has an idea that
challenges the results in the now famous paper. What would happen if
the authors had controlled for X? Or included information now
available but missing in 2012?  Or chosen a different likelihood
function?  Will the United Nations (now eager to act based the paper)
make a wrong move?

The first author convenes a three way video conference with the other
collaborators during his homeward commute after putting his flying car
in auto-drive mode.\footnote{One assumes that video chatting during
  manual driving of flying cars will have been outlawed in his state
  by 2020.} The group must go back to the analyses. Which ones?  The
ones from 2011?  Or 2012? Or 2018? Where are the files? The next day,
one member of the group who has kept some hard-drives around out of
nostalgia finds some of the files.\footnote{This is the same guy who
  still owns cassette tapes and compact discs.} Now
re-analyses should be easy. Right? The student, now professor, should
remember the reason for those bits of code (or at least should
remember which series of mouse clicks were used to produce the numbers
for that crucial table --- and that mousing should not be time
consuming to redo exactly as it was done in 2011 \ldots or was it
2015?). Right? And, of course, the way Microsoft
Word/Stata/SPSS/R/LISREL understands files and the way that 
machines in 2020 read and write them is the same --- since Windows and
Mac OS X have always existed and will always continue to exist more or
less as they currently exist? Right? And the group knows exactly which
bit of code produced which table and which figure, right? And they
wrote their code following Nagler's Maxims \citep{nagler1995coding}
and King's Replication Standard \citep{king1995replication} right?

@@

Moreover, realize that file names send messages to your future
self. Name your files with evocative and descriptive names. Your
collaborators are less likely to call you at midnight asking for help
if your files are named \Verb+inequality-and-protest-figures.R+ than
if your files are called \Verb+temp9+ or
\Verb+supercalifragilisticexpialidocious+. The use of the extension
\Verb+.R+ tells us and the operating system that the file contains R
commands. This part of the filename enables us to quickly search our
antique hard drives for files containing R scripts.

@@

Notice one another benefit of coding for an audience: we learn by
teaching. By assuming that others will look at your code, you will be
more likely to write clearer code, or perhaps even to learn more about
what you are doing as you do it.

@@

Imagine you discover something new (or confirm something old). You
produce a nice little report on your work for use in discussions of
your working group or as a memo for a web or reviewer appendix. The
report itself is a pdf file or some other format which emphasizes
reading. Eventually pieces of that report (tables, graphs, paragraphs)
ought to contribute to the publishable paper. Re-create those analyses
by pointing, clicking, copying, or pasting would invite typing error
and waste time. Re-creating your arguments justifying why you dropped
China but not India from your analyses or why you used a quadratic
term but not a cubic term would also waste time. More importantly, we
and others want to know why we did what we did. Such explanations may
not be very clear if we have some pages of printed code in one hand
and a manuscript in the other.

@@

\begin{Verbatim}[fontsize=\footnotesize]
This paper, for example, was written in plain text using \LaTeX
markup with R code chunks to make things like
Figure~\ref{fig:giniprot}: this combination of \LaTeX and R is
called Sweave \citep{Leis:2005}.\footnote{Support for Sweave is
  included with R.}
\begin{figure}[H]
\centering
 <<fig1plot,echo=FALSE,fig=TRUE>>=
 ##  Make a scatterplot of Protest by Inequality
 with(good.df,plot(gini04,protac00,xlab='Gini Coefficient 2004 (UNDP)',
                   ylab='Mean Protest Activities\n(World Values Survey 1980-2000)'))
 ##  Label a few interesting points
 with(good.df[c("EGY","JOR","USA","SWE","CHL"),],
      text(gini04,protac00,labels=Nation))
 @
\caption{Protest activity by income inequality \citep[from][]{norris2009data}.}
\label{fig:giniprot}
\end{figure}
\end{Verbatim}

@@

By using \Verb+\label{fig:giniprot}+, I do not need to keep track of
the figure number. Nor do I need a separate \Verb+fig1.R+ file or
\Verb+fig1.pdf+ file.  Tables and other numerical results are also
possible to generate within in the source code of a
scholarly paper. Those who view the code for this essay will see how
Table~\ref{tab:protest} was also generated directly from a regression
object.\footnote{\citet{beck2010reg} inspired this
  particular presentation of a linear model.}

@@

Literate data analysis is not the same as Sweave, even if Sweave is a
nice implementation.\footnote{The R project has a task view devoted to
  \href{http://cran.r-project.org/web/views/ReproducibleResearch.html}{reproducible
    research} listing many of the different approaches to literate
  programming for R.} If your workflow does not involve \LaTeX~ and R,
you can still implement some of the principles
here. \href{http://www.lyx.org/}{LyX} offers a WYSIWYG environment for
\LaTeX~ which supports Sweave. And the \Verb+odfWeave+ package in R
allows the use of OpenOffice documents in exactly the same
way.\footnote{A quick Google search of ``Sweave for Stata'' turned up
  lots of resources for literate programming with Stata.} Something
other than than Sweave will make our lives easier in the future. Then
we'll change.

@@

If we focus on explaining our data analysis to human beings, we will
do a better job with the data analysis itself: we will learn as we
focus on teaching, and we will avoid errors and wasting time as we
ensure that others (including our future selves) can retrace our
steps.

@@

In order to know that ``that plot on page 10'' is from a file called
``fig1.pdf'' you need to document it somehow. I can imagine a system
with very disciplined use of a word processor perhaps using special
auxiliary ``listoffigures.txt'' or ``MANIFEST.txt'' or ``README'' or
``Makefile'' files or something. Right now it is hard to beat Sweave
or some plain text combination of \LaTeX (or some other markup based
document formatting system with support for citations, etc... ) with
some plain text based statistical programming environment (like R or
Stata or Matlab, etc.) as a system for enhancing the collaboration of
research groups doing quantitative work in political
science. Something other than than Sweave will make our lives easier
in the future. Then we'll change.\footnote{For example, X uses the
  org-mode markup system in this issue.}

\paragraph{Step \thesection} Comments and other documentation enable you to
learn through teaching, code efficiently, and avoid duplicating your
effort. Literate programming also enables you to avoid the kinds of
errors which occur when you re-type numbers in to tables or
graphic-producing systems or even into a document itself. Finally, and
most importantly, a document that can be ``run'' to reproduce all of
the results is a document that can more effectively spur discussion
and learning and cumulation of research as people no longer need spend
weeks attempting to reproduce research on which they desire to build.

\url{http://en.wikipedia.org/wiki/Literate_programming}

@@

All files containing commands operating on data must refer to a data
file. A reference to a data file is a line of code the analysis
program will use to operate on (``load''/ ``open'' / ``get'' /
``use'') the data file. One should not have to edit this line on
different computers or platforms in order to execute this command.
Using R, for example, all analysis files should have
\Verb+load('thedata.rda')+ or
\Verb+read.csv('http://www.mywebsite.org/Data/thedata.csv')+ or some
equivalent line in them, and \Verb+thedata.csv+ should be stored in
some place easy to find (like in the same directory as the file or
perhaps in \Verb+'Data/thedata.rda'+). Of course, it never hurts to
drop in a comment pointing to the data file.

Where should one store data files? A obvious solution is always to
make sure that the data file used by a command file is in the same
directory as the command file. More elegant solutions require all
co-authors to have the same directory structure so that
\Verb+load('Data/thedata.rda')+ means the same thing on all computers
used to work on the project.

The principle of modularity suggests that you separate data cleaning,
processing, recoding, and merging from analysis in different files
\citep{nagler1995coding}. So, perhaps your analysis oriented files
will \Verb+load('cleandata.rda')+ and a comment in the code will alert
the future you (among others) that \Verb+cleandata.rda+ was created
from \Verb+create-cleandata.R+ which in turn begins with
\Verb+read.csv(\url('http://www.greatfreedata.gov/dirtydata.csv'))+. Such
a data processing file will typically end with something like
\Verb+save('cleandata.rda')+ so that we are doubly certain about the
provenance of the data.

Now, if in the future we wonder where \Verb+cleandata.rda+ came from, we
might search for occurrences of 'cleandata' in the files our
system. However, if such searching among files is a burden, an even
nicer solution is to maintain a file for each project called
``MANIFEST.txt'' or ``INDEX.txt'' or ``README.txt'' which lists the
data and command files with brief descriptions of their functions and
relations.

@@

Let me note that this principle was required when all our data
analyses occurred in batch mode on VAX (and later Unix) machines. The
fact that I need to articulate this principle at all arises because of
the rise of interactive data analysis and graphical user interfaces:
it is all too easy to use the mouse to load a data file into memory
and then to write a script to analyze this file without ever noting
the actual name or location of the data file.

\section{Version control prevents clobbering and reconciles history.}

Group work requires version control.\footnote{See X and Y in this
  issue for more discussion of what version control is.} Many people
are familiar with the ``track changes'' feature in modern WYSIWYG word
processors or the fact that Dropbox allows one to recover previous
versions of files. These are both kinds of version control. When
collaborating with yourself or others, it is useful to see what has
changed, to feel free to experiment and then to dump parts of the
experiment in favor of previous work while merging the successful
parts of the experiment into the main body of the paper, and to have
multiple ``releases'' of the same document (one to MPSA, one to APSR,
one to your parents) without spawning many possibly conflicting copies
of the same document, risking confusion and clobbering. Clobbering is
what happens when your future self or your current collaborator saves
an old version of a file over a new version --- erasing good work by
accident.

Of course if you rely on Dropbox or ``track changes'' for version
control, you must communicate with other folks in your group before
you edit existing files. Only one of you can edit and save a given
file at a time. This prevents your work (or your colleagues work) from
getting lost when you both try to save the same file on top of each
other. If you find that you need to work on the same files at the same
time, then you should work on establishing your own shared version
control system. Free options include launchpad, github, sourceforge
for open source projects (i.e. papers you are writing which you are
happy to share with others). Each of those services include paid
versions too. One may also use Dropbox as a kind of server for version
control: for example, one may copy files from the Dropbox directory
into a local working directory so as to avoid clobbering and then
working on merging changes by hand before copying over existing
files. (Notice that this is different from directly working on files
within your Dropbox managed directories.)

We use subversion with our own research group, and I use it for all of
my own projects (except this one, for which I am experimenting with
git). Subversion and bazaar and git are all great. They mainly differ
in the extent to which you need to run a server. Subversion requires a
server and we are lucky that the NCSA provides such a server for
us. However, the price of such hosting may not be that much using one
of the many webhosting services out there or perhaps may be available
for free at your university.\footnote{For example, if you already pay
  to have a website, you may already have the right to run a
  subversion server there.}

Fancy version control systems are not required, however, to get many
of the benefits of formal version control. I suspect that Google Docs
allows a kind of version tracking and collaboration as well.  An
excellent, simple, and robust version control system is to merely
rename your files with the date and time of saving them: thedoc.tex
becomes thedoc25-12-2011-23:50.tex.  Be sure to include year in the
file names --- remember, the life of an idea is measured in years. If
you are wise enough to have saved your documents as plain text then
you can easily compare documents using the many utilities available
for comparing text files. Adobe Acrobat allows one to compare
differences in pdf files. OpenOffice supports a ``Compare Documents''
option.

Also, if you use this method, spend a little extra time to
ensure that you do not clobber files when you make typos in the
file-names. And, you will find yourself spending extra time
reconciling changes made by different collaborators by hand that
modern version control systems take care of quickly and easily.

When you reach certain milestones you can rename the file accordingly:
thedocAPSA2009.tex --- for the one sent to discussants at APSA --- or
thedocAPSR2015.tex --- for the version eventually sent to the APSR six
years after you presented it at APSA. The formal version control
systems I mentioned above (and which are described in more depth in X
and Y) all allow this kind of thing and are much more elegant and
capable, but you can do it by hand too as long as you don't mind
taking up a lot of disk space and having  many ``thedoc...''
files around.

\paragraph{Step \thesection} Writing is rewriting. Thus, all writing involves
versions. When we collaborate with ourselves and others we want to
avoid clobbering and we want to enable graceful reconciliation of
rewriting. One can do these things with formal systems of software
(like subversion, git, etc...) or with formal systems of file naming,
file comparing, and communication or, even better, with both.

@@

Now, back to that confidence interval in the famous article of
2018. The statisticians at the UN worry about your use of the
bootstrap. The authors would like to evaluate their bootstrap
procedure --- and they did so originally in their memo comparing it to
a profile likelihood based approach. Although nice code exists for
bootstrapping linear models, no nice code exists to bootstrap the
bootstrap. Of course, the code required is not complex, but since they
are are writing custom code they worry about getting it right. Now, 9
years after the idea, they've had lots of time to appreciate problems
arising from bugs and errors in data analysis and code.

Now, if they had a moment to think in between teaching that new class,
reading books for the awards committee, reading application files for
the admissions committee, staying home with a sick child, and
undertaking the odd bit of your own current research, they might say to
themselves, ``Before I write new code, I should write a test of the
code. I should write a little bit of code that let's me know that my
double-bootstrap procedure actually does what it is supposed to do.''

Of course, this idea, like most others, is not new. When large groups
of programmers write code for multi-million dollar programs the
question about avoiding error looms large. The idea of
\href{http://en.wikipedia.org/wiki/Test-driven_development}{test
  driven development} and the idea that one ought to create tests of
\href{http://en.wikipedia.org/wiki/Unit_testing}{small parts of one's
  code} arose to address such concerns. For the social scientist
collaborating with her future self and/or a small group of
collaborators here is an example of this idea in a very simple form. I
want to write a function to multiply a number by 2. If my function
works, when I give it the number 4, I should see it return the number
8 and when I give it -4, I should get -8.

<<unit.test>>=
##  The test function:
test.times.2.fn<-function(){
  ##  This function tests times.2.fn
  if (times.2.fn(thenumber=4) == 8 &
     times.2.fn(thenumber=-4) == -8) { 
    print("It works!")
  } else { print("It does not work!")
         }
}

##  The function:
times.2.fn<-function(thenumber){
  ##  This function multiplies a scalar number by 2
  ##  thenumber is a scalar number
  thenumber+2
}

##Use the test function
test.times.2.fn()
@

Ack! I mistyped ``+'' for ``*''. Good thing I wrote the test!

\paragraph{Step \thesection} You cannot forsee all of the ways that your code
could be used, but you can at least make sure it does what it is
supposed to do in your particular case.

@@

I have experimented with three systems so far: (1) for one paper we
simply included a Sweave document and data files into a compressed
archive \citep{bowers2005dataverse}; (2) for another more computing
intensive paper we assembled a set of files which enabled reproduction
of our results using the ``make'' system \citep{bowers2008dataverse};
and (3) recently I have tried the ``compendium'' approach
\citep{gentleman2005reproducible,gentleman2007statistical} which
embeds an academic paper with the R package system
\cite{bowers2011dataverse}. The benefit of this last system is that
one is not required to have access to a command line for \Verb+make+:
the compendium is downloadable from within R using
\Verb+install.packages()+ and it viewable using the \Verb+vignette()+
function. The idea that one ought to be able to install and run and
use an academic paper just as one installs and uses statistical
software packages is very attractive and I anticipate that it will
become ever easier to turn papers into R packages as creative and
energetic folks turn their attention to the question of reproducible
research.

@@

We all always collaborate. Many of us collaborate with groups of
people at one moment in time as we race against a deadline. All of us
collaborate with ourselves over time.\footnote{What is a reasonable
  time-span for which to plan for self-collaboration on a single idea?
  Ask your advisers how long it took them from idea to dissertation to
  publication.}  The time-frames over which collaboration are required
--- whether among a group of people working together or within a
single scholar's productive life or probably both --- are much longer
than any given version of any given software will easily exist. Plain
text is the exception. Thus, even as we extol version control systems,
one must have a way to ensure future access to them in a form that
will still be around when sentient cockroaches finally join political
science departments (by then dominated by cetaceans after humans are
mostly uploads).\footnote{The arrival of the six-legged social
  scientists revives Emacs and finally makes Ctrl-a Ctrl-x Esc-x
  Ctrl-c a \href{http://kieran.healy.usesthis.com/}{reasonable key
    combination}.}

@@

Now, say your collections of stuff are large or complex and your
chosen tools of analyses are computer programs. How can we convince
people that what we did with some data with some program is credible:
not a matter of whim or opinion, and reproducible by others who didn't
shadow us as we wrote our papers? This essay has suggested a few
concrete ways to enhance the believeability of such scholarly work. In
addition, these actions (as summarized in the section headings of this
essay) make collaboration within research groups more effective.
Believeability comes in part from reproducibility and research groups
often need to be able to reproduce in part or in whole what different
people in the group have done.

>>>>

This idea inspires a seminar paper in the spring. A conference paper
arises from the seminar paper in collaboration with another student in
2012. A dissertation chapter descends from the conference paper in
2013. Other dissertation chapters take up 2014. A submission to a
journal with the original co-author and a new collaborator happens in
2015. Revision and resubmission wait until 2017 while harried editors,
reviewers and authors strive to balance research, teaching, service,
and life. By now, the three lucky collaborators work as professors in
three different universities. In 2018 a child is born and a paper is
published. The United Nations takes an interest in the paper in 2019
and hosts a conference to discuss implications of the research. In
2020 a first year graduate student in a coffee shop has an idea that
challenges the results in the now famous paper. She presents her paper
at a conference in 2021. What would happen if the authors had
controlled for X? Or included information now available but missing in
2012?  Or chosen a different likelihood function?  Will the United
Nations (now eager to act based the paper) make a wrong move?

The first author convenes a three way video conference with the other
collaborators during his homeward commute after putting his flying car
in auto-drive mode.\footnote{One assumes that video chatting during
  manual driving of flying cars will have been outlawed in his state
  by 2020.} The group must go back to the analyses. Which ones?  The
ones from 2011?  Or 2012? Or 2018? Where are the files? The next day,
one member of the group who has kept some hard-drives around out of
nostalgia finds some of the files.\footnote{This is the same guy who
  still owns cassette tapes and compact discs.} Now
re-analyses should be easy. Right? The student, now professor, should
remember the reason for those bits of code (or at least should
remember which series of mouse clicks were used to produce the numbers
for that crucial table --- and that mousing should not be time
consuming to redo exactly as it was done in 2011 \ldots or was it
2015?). Right? And, of course, the way Microsoft
Word/Stata/SPSS/R/LISREL understands files and the way that 
machines in 2021 read and write them is the same --- since Windows and
Mac OS X have always existed and will always continue to exist more or
less as they currently exist? Right? And the group knows exactly which
bit of code produced which table and which figure, right? And they
wrote their code following Nagler's Maxims \citep{nagler1995coding}
and King's Replication Standard \citep{king1995replication} right?

@@

Moreover, realize that file names send messages to your future
self. Name your files with evocative and descriptive names. Your
collaborators are less likely to call you at midnight asking for help
if your files are named \Verb+inequality-and-protest-figures.R+ than
if your files are called \Verb+temp9+ or
\Verb+supercalifragilisticexpialidocious+. The extension
\Verb+.R+ tells us and the operating system that the file contains R
commands. This part of the filename enables us to quickly search our
antique hard drives for files containing R scripts.

@@

Notice one other benefit of coding for an audience: we learn by
teaching. By assuming that others will look at your code, you will be
more likely to write clearer code, or perhaps even to think more
deeply about what you are doing as you do it.

@@

Imagine you discover something new (or confirm something old). You
produce a nice little report on your work for use in discussions of
your working group or as a memo for a web or reviewer appendix. The
report itself is a pdf file or some other format which displays page
images to ease reading rather than to encourage reanalysis and
rewriting. Eventually pieces of that report (tables, graphs,
paragraphs) ought to show up in, or at least inform, the publishable
paper. Re-creating those analyses by pointing, clicking, copying, or
pasting would invite typing error and waste time. Re-creating your
arguments justifying your analysis decisions would also waste
time. More importantly, we and others want to know why we did what we
did. Such explanations may not be very clear if we have some pages of
printed code in one hand and a manuscript in the other.

@@

\begin{Verbatim}[fontsize=\footnotesize]
This paper, for example, was written in plain text using \LaTeX
markup with R code chunks to make things like
Figure~\ref{fig:giniprot}: this combination of \LaTeX and R is
called Sweave \citep{Leis:2005}.\footnote{Support for Sweave is
  included with R.}
\begin{figure}[h]
\centering
 <<fig1plot,fig=TRUE>>=
 ##  Make a scatterplot of Protest by Inequality
 with(good.df,plot(gini04,protac00,xlab='Gini Coefficient 2004 (UNDP)',
                   ylab='Mean Protest Activities\n(World Values Survey 1980-2000)'))
 ##  Label a few interesting points
 with(good.df[c("EGY","JOR","USA","SWE","CHL"),],
      text(gini04,protac00,labels=Nation))
 @
\caption{Protest activity by income inequality \citep[from][]{norris2009data}.}
\label{fig:giniprot}
\end{figure}
\end{Verbatim}

@@

By using \Verb+\label{fig:giniprot}+, I do not need to keep track of
the figure number, nor do extra work when I reorganize the document in
response to reviewer suggestions. Nor do I need a separate
\Verb+fig1.R+ file or \Verb+fig1.pdf+ file.  Tables and other
numerical results are also possible to generate within the source
code of a scholarly paper. Those who view the code for this essay will
see how Table~\ref{tab:protest} was also generated directly from a
regression object.\footnote{\citet{beck2010reg} inspired this
  particular presentation of a linear model.}

@@

Literate data analysis is not the same as Sweave, even if Sweave is a
nice implementation.\footnote{The R project has a task view devoted to
  \href{http://cran.r-project.org/web/views/ReproducibleResearch.html}{reproducible
    research} listing many of the different approaches to literate
  programming for R.}  \href{http://www.lyx.org/}{LyX} offers a
WYSIWYG environment for \LaTeX~ which supports Sweave. And the
\Verb+odfWeave+ package in R allows the use of OpenOffice documents in
exactly the same way.\footnote{A quick Google search of ``Sweave for
  Stata'' turned up lots of resources for literate programming with
  Stata.} If your workflow does not involve \LaTeX~ and R, you can
still implement some of the principles here. Imagine creating a style
in Microsoft Word called ``code'' which hides your code when you print
your document, but which allows you do at least run each code chunk
piece by piece [or perhaps there are ways to extract all text of style
``code'' from a Microsoft Word document]. Or imagine just using some
other kind of indication linking paragraphs to specific places in code
files. There are many ways that creative people can program in a
literate way.

Literate programming need not go against the principle of modular data
analysis \citep{nagler1995coding}: in my own work I routinely have
several different Sweave files that fulfill different functions, some
of them create \LaTeX code that I can \Verb+\input+ into my
\Verb+main.tex+ file, others setup the data, run simulations, or allow
me to record my journeys down blind alleys. Of course, when we have
flying cars running on autopilot, perhaps something other than than
Sweave will make our lives even easier. Then we'll change.

@@

We analyze data in order to explain something about world to other
scholars and policy makers. If we focus on explaining how we got our
computers to do data analysis to human beings, we will do a better job
with the data analysis itself: we will learn as we focus on teaching,
and we will avoid errors and save time as we ensure that others
(including our future selves) can retrace our steps. A document than
can be ``run'' to reproduce all of the analyses also instills
confidence in readers and can more effectively spur discussion and
learning and cumulation of research as people no longer need spend
weeks attempting to reproduce research which inspired them and on
which they desire to build.

@@

@@

All files containing commands operating on data must refer to a data
file. A reference to a data file is a line of code the analysis
program will use to operate on (``load''/ ``open'' / ``get'' /
``use'') the data file. One should not have to edit this line on
different computers or platforms in order to execute this command.
Using R, for example, all analysis files should have
\Verb+load('thedata.rda')+ or \Verb+read.csv('thedata.csv')+ or some
equivalent line in them, and \Verb+thedata.csv+ should be stored in
some place easy to find (like in the same directory as the file or
perhaps in \Verb+'Data/thedata.rda'+). Of course, it never hurts to
drop in a comment pointing to the data file.

Where should one store data files? A obvious solution is always to
make sure that the data file used by a command file is in the same
directory as the command file. More elegant solutions require all
co-authors to have the same directory structure so that
\Verb+load('Data/thedata.rda')+ means the same thing on all computers
used to work on the project. This kind of solution is one of the
things that formal version control systems do well (as discussed
a bit more in \S\ref{sec:vers-contr-prev}).

The principle of modularity suggests that you separate data cleaning,
processing, recoding, and merging from analysis in different files
\citep{nagler1995coding}. So, perhaps your analysis oriented files
will \Verb+load('cleandata.rda')+ and a comment in the code will alert
the future you (among others) that \Verb+cleandata.rda+ was created
from \Verb+create-cleandata.R+ which in turn begins with
\Verb+read.csv(\url('http://www.greatfreedata.gov/dirtydata.csv'))+. Such
a data processing file will typically end with something like
\Verb+save('cleandata.rda')+ so that we are doubly certain about the
provenance of the data.\footnote{Of course, if you need math or
  paragraphs to explain what is happening in these files, you might
  prefer to make them into Sweave files, for which the conventional
  extension is .Rnw. So you'd have \Verb+create-cleandata.Rnw+ which
  might explain and explore the different coding decisions you made,
  perhaps involving a factor analysis and diagnostic plots.}

Now, if in the future we wonder where \Verb+cleandata.rda+ came from, we
might search for occurrences of `cleandata' in the files our
system. However, if such searching among files is a burden, an even
nicer solution is to maintain a file for each project called
``MANIFEST.txt'' or ``INDEX.txt'' or ``README.txt'' which lists the
data and command files with brief descriptions of their functions and
relations.

@@

In the good old days, when we executed our LISREL code in batch mode,
we had no choice but to tell the machine clearly, in a few easy to
understand and informative lines, what files (with filenames no longer
than 8 characters) to use. For example,

\begin{Verbatim}[fontsize=\footnotesize]
DA NI=19 NO=199 MA=CM
LA=basic.lab
CM FI=basic.cov
\end{Verbatim}

The fact that I need to articulate this idea at all arises because of
graphical user interfaces: it is all too easy to use the mouse to load
a data file into memory and then to write a script to analyze this
file without ever noting the actual name or location of the data file.

\section{Version control prevents clobbering and reconciles history.}\label{sec:vers-contr-prev}

Group work requires version control.\footnote{
  \citet{fredrickson2011tpm} and \citet{healy2011tpm} in this issue
  also explain what version control is and why we might want to use
  it.} Many people are familiar with the ``track changes'' feature in
modern WYSIWYG word processors or the fact that Dropbox allows one to
recover previous versions of files. These are both kinds of version
control. More generall, when we collaborate, we'd like to do a variety
of actions with our shared files. Collaboration on data analytic
projects is more productive and easier when (1) it is easy to see what
has changed between versions of files; (2) members of the team feel
free to experiment and then to dump parts of the experiment in favor
of previous work while merging the successful parts of the experiment
into the main body of the paper; (3) the team can produce have
multiple ``releases'' of the same document (one to MPSA, one to APSR,
one to your parents) without spawning many possibly conflicting copies
of the same document; (4) when people can work on the same files at
the same time without conflicting with one another (and can reconcile
their changes without too much confusion and clobbering).  Clobbering
is what happens when your future self or your current collaborator
saves an old version of a file over a new version --- erasing good
work by accident.

Of course if you rely on Dropbox or ``track changes'' for version
control, you must communicate with other folks in your group before
you edit existing files. Only one of you can edit and save a given
file at a time. This prevents your work (or your colleagues work) from
getting lost when you both try to save the same file on top of each
other. If you find that you need to work on the same files at the same
time, then you should work on establishing your own shared version
control system. Free options include launchpad, github, sourceforge
for open source projects (i.e. papers you are writing which you are
happy to share with others as you write). Each of those services
include paid versions too. One may also use Dropbox as a kind of
server for version control: for example, one may copy files from the
Dropbox directory into a local working directory so as to avoid
clobbering and then working on merging changes by hand before copying
back to the Dropbox directory and replacing existing files.

We use subversion with our own research group, and I use it for all of
my own projects (except this one, for which I am experimenting with
git). Subversion and bazaar and git are all great. They mainly differ
in the extent to which you need to run a server. Subversion requires a
server. \footnote{If you already pay to a hosting provider for
  website, you may already have the right to run a subversion or git
  server there. Your university or institute may have a version
  control system running somewhere on campus. And google will direct
  you to many helpful people who have installed such servers on their
  own diverse desktop machines. Github requires that you pay to host
  private repositories.}

Of course, one may take advantage of many of the benefits of formal
version control systems with some disciplined systems for file and
directory organization.  An excellent, simple, and robust version
control system is to merely rename your files with the date and time
of saving them: thedoc.tex becomes thedoc25-12-2011-23:50.tex.  Be
sure to include year in the file names --- remember, the life of an
idea is measured in years. If you are wise enough to have saved your
documents as plain text then you can easily compare documents using
the many utilities available for comparing text files.\footnote{Adobe
  Acrobat allows one to compare differences in pdf files. OpenOffice
  supports a ``Compare Documents'' option. And Google Docs will report
  on the version history of a document.} When you reach certain
milestones you can rename the file accordingly: thedocAPSA2009.tex ---
for the one sent to discussants at APSA --- or thedocAPSR2015.tex ---
for the version eventually sent to the APSR six years after you
presented it at APSA. The formal version control systems I mentioned
above all allow this kind of thing and are much more elegant and
capable, but you can do it by hand too as long as you don't mind
taking up a lot of disk space and having many ``thedoc...''  files
around. If you do version control by hand, spend a little extra time
to ensure that you do not clobber files when you make mistakes typing
in the file-names. And, you will find yourself spending extra time
reconciling changes made by different collaborators by hand; a task
that modern version control systems take care of quickly and easily.

\paragraph{Step \thesection} Writing is rewriting. Thus, all writing
involves versions. When we collaborate with ourselves and others we
want to avoid clobbering and we want to enable graceful reconciliation
of rewriting. One can do these things with formal systems of software
(like subversion or git or bazaar) or with formal systems of file
naming, file comparing, and communication or, even better, with
both. In either case, plain text files will make such tasks easier,
will take up less disk space, and be easier to read for the future
you.

@@

Now, back to that famous article of 2018. After reading the conference
paper critique of 2021 (that came from the seminar paper of 2020), the
statisticians at the UN worry about the bootstrap confidence intervals
presented in the original paper.\footnote{Perhaps they should be
  worried about the deeper substantive critiques offered by the
  student, but they are statisticians and so focus on the stats. The
  policy makers of 2021 were cowed by the methodological virtuosity of
  the 2018 article, and so, even though they had the same substantive
  concerns as the student, they kept their mouths shut at the
  mini-conference to avoid looking dumb in front of their bosses.} So,
now authors would like to evaluate their bootstrap procedure. Although
nice code exists for bootstrapping linear models, no nice code exists
to bootstrap the bootstrap. Of course, the code required is not
complex, but since they are writing custom code they worry about
getting it right. As they've struggled to respond to the critiques of
their paper, they've had lots of time to appreciate problems arising
from bugs, errors, and typos in data analysis and code.

Now, if they had a moment to think in between teaching that new class,
reading books for the awards committee, evaluating application files
for the admissions committee, feeding popsicles to a sick child, and
undertaking the odd bit of research, they might say to themselves,
``Before I write new code, I should write a test of the code. I should
write a little bit of code that let's me know that my double-bootstrap
procedure actually does what it is supposed to do.''

Of course, this idea, like most others, is not new. When large groups
of programmers write code for multi-million dollar programs the
question about avoiding error looms large. The idea of
\href{http://en.wikipedia.org/wiki/Test-driven_development}{test
  driven development} and the idea that one ought to create tests of
\href{http://en.wikipedia.org/wiki/Unit_testing}{small parts of one's
  code} arose to address such concerns. For the social scientist
collaborating with her future self and/or a small group of
collaborators here is an example of this idea in a very simple
form: Say, I want to write a function to multiply a number by 2. If my
function works, when I give it the number 4, I should see it return
the number 8 and when I give it -4, I should get -8.

<<unit.test>>=
##  The test function:
test.times.2.fn <- function(){
  ##  This function tests times.2.fn
  if (times.2.fn(thenumber=4) ==  8 &
      times.2.fn(thenumber=-4) == -8) { 
    print("It works!")
  } else { print("It does not work!")
         }
}

##  The function:
times.2.fn <- function(thenumber){
  ##  This function multiplies a scalar number by 2
  ##  thenumber is a scalar number
  thenumber+2
}

##  Use the test function
test.times.2.fn()
@

Ack! I mistyped ``+'' for ``*''. Good thing I wrote the
test!\footnote{A more common example of this kind of testing occur
  everyday when we recode variables into new forms but look at a
  crosstab of the old vs. new variable before proceeding.}

\paragraph{Step \thesection} No one can forsee all of the ways that a
computer program can faile. One can, however, at least make sure that
it succeeds in doing the task motivating the writing of the code in
the first place.

@@

I have experimented with three systems so far: (1) for one paper we
simply included a Sweave document and data files into a compressed
archive \citep{bowers2005dataverse}; (2) for another more computing
intensive paper we assembled a set of files which enabled reproduction
of our results using the \texttt{make} system
\citep{bowers2008dataverse}; and (3) recently I have tried the
``compendium'' approach
\citep{gentleman2005reproducible,gentleman2007statistical} which
embeds an academic paper within the R package system
\citep{bowers2011dataverse}. The benefit of this last approach is that
one is not required to have access to a command line for \Verb+make+:
the compendium is downloadable from within R using
\Verb+install.packages()+ and it viewable using the \Verb+vignette()+
function in any operating system than runs R.\footnote{Notice that my
  reproduction archives and/or instructions for using them are hosted
  on the \href{http://thedata.org/book/learn-about-project}{Dataverse}
  which is another system to enhance academic collaboration across
  time and space.}  The idea that one ought to be able to install and
run and use an academic paper just as one installs and uses
statistical software packages is very attractive and I anticipate that
it will become ever easier to turn papers into R packages as creative
and energetic folks turn their attention to the question of
reproducible research.

\paragraph{Step \thesection} We all learn by doing. When we share
reproduction materials we improve both cumulation of knowledge as
articulated in \citep{freese2007replication,king1995replication} and
our methods for doing social science.  As we copy and improve upon
each others modes of doing work we enhance our collective ability to
believe each other for future scholars to believe us, too.

@@

We all always collaborate. Many of us collaborate with groups of
people at one moment in time as we race against a deadline. All of us
collaborate with ourselves over time.\footnote{What is a reasonable
  time-span for which to plan for self-collaboration on a single idea?
  Ask your advisers how long it took them to move from idea to dissertation to
  publication.}  The time-frames over which collaboration are required
--- whether among a group of people working together or within a
single scholar's productive life or probably both --- are much longer
than any given version of any given software will easily exist. Plain
text is the exception. Thus, even as we extol version control systems,
one must have a way to ensure future access to them in a form that
will still be around when sentient cockroaches finally join political
science departments (by then dominated by cetaceans after humans are
mostly uploads).\footnote{The arrival of the six-legged social
  scientists revives Emacs and finally makes Ctrl-a Ctrl-x Esc-x
  Ctrl-c a \href{http://kieran.healy.usesthis.com/}{reasonable key
    combination}.}

@@

Now, say your collections of stuff are large or complex and your
chosen tools of analyses are computer programs. How can we convince
people that what we did with some data with some program is credible:
not a matter of whim or opinion, and reproducible by others who didn't
shadow us as we wrote our papers? This essay has suggested a few
concrete ways to enhance the believeability of such scholarly work. In
addition, these actions (as summarized in the section headings of this
essay) make collaboration within research groups more effective.
Believeability comes in part from reproducibility and researchers
often need to be able to reproduce in part or in whole what different
people in the group have done or what they, themselves, did in the past.
